---
title: "A Data visualisation project : Exploratory Data Analysis for SARS-COV-2 seropositivity"
author: "Audrey T Yeo, M Sci Biostats, M Nursing, RN, audreytyeo@gmail.com"
output: pdf_document
---
> Document was recreated in a reproducible format

## Loading relevant packages
<!-- Loading relevant packages and data -->
```{r, echo = TRUE, error = FALSE, message = FALSE, warning= FALSE}
library(readxl)
library(ggplot2)
library(janitor)
library(kableExtra)
library(tidyverse)
df = read.csv("serocov-pop_data_public.txt", sep = ",")
```
## Data engineering
```{r}
set.seed(521)
names(df) #names of variables, there are 11 variables
dim(df) # 1000 x 11
# data engineering
df$pos[df$pos == 1] = "positive"
df$pos[df$pos == 0] = "negative"
df$Sex[df$Sex == "1"] = "male"
df$Sex[df$Sex == "0"] = "female"
df$age = as.numeric(df$age)
names(df)[8] <- paste("FinalResult")
df$bmi = as.vector(sample(19:51, 1000, replace = TRUE))
n = dim(df)[1]
a = .13*n
b = .15*n
c = .10*n
d = .20*n
e = .03*n
f = .04*n
g = .01*n
h = .34*n
df$comorbidities = c(rep("acute respiratory", a), 
                     rep("chronic respiratory", b), rep("cancer", c), 
                     rep("diabetes", d), rep("sedentary", e), 
                     rep("other endocronology", f), rep("chronic renal", g),
                     rep("immunosupression", h))
```

<!-- Data is reshaped to long form and some variables are recoded -->
```{r, echo = FALSE, error = FALSE, message = FALSE, warning= FALSE}
# Data frame for facet change for comodbidities

```

#Aim

The aim of the exploratory data analysis is to analyse the major risk factors for an infectious disease diagnosed bySARS-COV-2 seropositivy *based on synthetic data*

## Descriptives on data

There are `r dim(df)[1]` individuals in this study and `r length(df)` variables. There are `r length(df$FinalResult[df$FinalResult == "positive"])` cases and `r length(df$FinalResult[df$FinalResult == "negative"])` controls. Cases are defined as SARS-COV-2 seropositive denoted by outcome "positive" in variable "FinalResult". Across `r length(df)` variables, there are seven categories of comorbidities which can be understood as risk factors. Overall there are only `r sum(is.na(df$FinalResult))` missing values for these categories. The cases represented approximately by the incidence of `r round(dim(df$FinalResult[df$FinalResult == "positive"])[1]/dim(df)[1],2)*100`% of the study population. The breakdown of sexes are seen in the figure below. There is only one subject in a third sex group and this was a negative test. With respect to distribution in percentages, there is an equal proportion of male and female sexes within the cases and control group. 
A careful approach would be needed to impute and infer based on the available data for these variables.

```{r counts, echo = FALSE, error = FALSE, message = FALSE, warning= FALSE}

# frequency table of cases
df %>% tabyl(FinalResult, Sex) %>% 
  adorn_rounding(digits = 2) %>% 
  adorn_totals(where = c("row", "col")) -> counts

colnames(counts) = c("Final result", "male", "female", "total")

knitr::kable(counts, row.names = FALSE, booktabs = FALSE, format = "latex", align = "c", longtable = TRUE)

```

# Gender

## Cases and controls by gender

The proportions of cases and controls by gender are as follows. Proportions are the same in each sex group for each final result.

```{r, echo = FALSE, error = FALSE, message = FALSE, warning= FALSE}
df %>% tabyl(FinalResult) %>% 
  adorn_percentages(denominator = "col") %>% 
  adorn_rounding(digits = 2)-> fre

colnames(fre) = c("Final result", "count", "percent")

knitr::kable(fre, row.names = FALSE, booktabs = FALSE, format = "latex", align = "c", longtable = TRUE)
```
\newpage
# Age

## Cases and controls by age and gender

Age across cases is (n = `r dim(df[df$FinalResult == "positive",])[1]`) and controls (n = `r dim(df[df$FinalResult == "negative",])[1]`). For the cases, males are on average `r round(mean(df$age[df$FinalResult == "positive" & df$Sex == "male"], na.rm =TRUE),2)` years old. The difference between male and female ages in cases are approximately `r round(mean(df$age[df$FinalResult == "positive" & df$Sex == "male"], na.rm =TRUE),2)-round(mean(df$age[df$FinalResult == "positive" & df$Sex == "female"], na.rm =TRUE),2)` years younger than their female counterparts. Furthermore, the distribution of age across cases and controls are are `r round(mean(df[df$FinalResult == "positive",]$age),2)` [`r min(df[df$FinalResult == "positive",]$age, na.rm = TRUE)`, `r max(df[df$FinalResult == "positive",]$age)`] and controls `r round(mean(df[df$FinalResult == "negative",]$age), 2)` [`r min(df[df$FinalResult == "negative",]$age)`, `r max(df[df$FinalResult == "negative",]$age)`]. 

## Cases by age and sex

The median age below and above 50 are `r median(df[df$age < 50 & df$Sex == "f",]$age)` for females and `r median(df[df$age < 50 & df$Sex == "f",]$age)` for males. 

```{r , echo = FALSE, error = FALSE, message = FALSE, warning= FALSE, fig.align = "c", fig.height=4, fig.width=8}
ggplot(df[df$FinalResult=="positive",], aes(x = age, fill = Sex)) + 
  geom_density(alpha = 0.5, colour = "lightblue") + theme_gray() +
  labs(x = "age (years)", title = "Age distribution of cases for each sex with likewise-coloured \n means indicated by vertical lines", fill = "Sex") +  
  geom_vline(xintercept = round(mean(df$age[df$Sex == "male"]), 2), colour = "red", linetype = "dotted", pch = 5) +
  geom_vline(xintercept = round(mean(df$age[df$Sex == "female"]), 2), colour = "darkgreen", linetype = "dotted", pch = 3) +
  scale_x_continuous(breaks = c(seq(4,92,4)))
```

## Controls by age and sex

The median age below and above 50 are `r median(df[df$age < 50 & df$FinalResult == "negative",]$age)` for females and `r median(df[df$age < 50 & df$FinalResult == "positive",]$age)` for males. 

```{r, echo = FALSE, error = FALSE, message = FALSE, warning= FALSE, fig.height=4, fig.width=8, fig.align="c"}
ggplot(df[df$FinalResult=="negative",], aes(x = age, fill = Sex)) + 
  geom_density(alpha = 0.5, colour = "lightblue") + theme_gray() +
  labs(x = "age (years)", title = "Age distribution of controls for each sex with likewise-coloured \n means indicated by vertical lines", fill = "Sex") +  
  geom_vline(xintercept = 52.01, colour = "red", linetype = "dotted", pch = 5) +
  geom_vline(xintercept = 51.48, colour = "darkgreen", linetype = "dotted", pch = 3) +
  geom_vline(xintercept = 68, colour = "lightblue", linetype = "dotted", pch = 10) +
  scale_x_continuous(breaks = c(seq(4,92,4)))

# print(summary(df$age[df$FinalResult == "negative" & df$Sex == "m"]))
# print(summary(df$age[df$FinalResult == "negative" & df$Sex == "n"]))
```

## Cases and controls by age

Cases and controls are similarly distributed by age. 

```{r, echo = FALSE, error = FALSE, message = FALSE, warning= FALSE, fig.height=4, fig.width=8, fig.align="c"}
ggplot(df, aes(x = age, fill = FinalResult)) + 
  geom_density(alpha = 0.5, colour = "lightblue") + theme_gray() +
  labs(x = "age (years)", title = "Age distribution of cases and controls with likewise-coloured \n means indicated by vertical lines", fill = "Final result")  +
  geom_vline(xintercept = 51.73, colour = "pink", linetype = "dotted", pch = 3) +
  geom_vline(xintercept = 46.69, colour = "lightblue", linetype = "dotted", pch = 10) +
  scale_x_continuous(breaks = c(seq(4,94,4)))

# summary(df$age[df$FinalResult == "negative"])
# summary(df$age[df$FinalResult == "positive"])
```

\newpage
## Cases and controls by age in raw counts

Raw counts and distribution is seen below of age distribution in cases and controls. 

```{r, echo = FALSE, error = FALSE, message = FALSE, warning= FALSE,fig.height=4, fig.width=8, fig.align="c"}
# Age distribution amongst positive cases
ggplot(df) +
    geom_histogram(aes(x= age, fill=FinalResult,
                       position = 'identity'))  +
  theme_gray() + labs(x = "age (years)", y = "count", fill = "Final Result") +
  scale_x_continuous(breaks = c(seq(4, 96, 4))) + labs(title = "Cases and controls by age in counts") +
  geom_vline(xintercept = mean(df$age[df$Sex == "female"]), colour = "red", linetype = "dotted", pch = 5) +
  geom_vline(xintercept = mean(df$age[df$Sex == "male"]), colour = "darkgreen", linetype = "dotted", pch = 3) 

#summary(df[df$FinalResult == "positive",]$age)[4]

#summary(df[df$FinalResult == "negative",]$age)
```

## Cases by sex and age



```{r, echo = FALSE, error = FALSE, message = FALSE, warning= FALSE,fig.height=4, fig.width=8, fig.align="c"}
# Age distribution amongst neg cases
ggplot(df[df$FinalResult == "positive",]) +
    geom_histogram(aes(x= age, fill=Sex,
                       position = 'identity'))  +
  theme_gray() + labs(x = "age (years)", y = "count", fill = "sex") +
  scale_x_continuous(breaks = c(seq(18, 96, 4))) + labs(title = "Sex and age for cases in counts") +
  geom_vline(xintercept = mean(df$age[df$Sex == "female"]), colour = "red", linetype = "dotted", pch = 5) +
  geom_vline(xintercept = mean(df$age[df$Sex == "male"]), colour = "darkgreen", linetype = "dotted", pch = 3) 

```

#BMI


```{r, echo = FALSE, error = FALSE, message = FALSE, warning= FALSE}

tabyl(df$bmi) %>% 
  adorn_rounding() -> bmi

#bmi = bmi[40:50, 1:2]

# colnames(bmi) = c("bmi value", "count")
# 
# knitr::kable(bmi, row.names = FALSE, booktabs = FALSE, format = "latex", align = "c", longtable = TRUE)

```

##BMI across cases and controls

```{r, echo = FALSE, error = FALSE, message = FALSE, warning= FALSE,fig.height=4, fig.width=8, fig.align="c"}

ggplot(df[df$bmi < 41,], aes(x = bmi, fill = FinalResult)) + 
  geom_density(alpha = 0.5, colour = "lightblue") + theme_gray() +
  labs(x = "bmi", title = "Distribution of bmi across cases and controls", fill = "Final result") 

```
\newpage

## Comorbidity  in cases

```{r, echo = FALSE, error = FALSE, message = FALSE, warning= FALSE,fig.height=4, fig.width=8, fig.align="c"}
ggplot(df, aes(x = age, fill = comorbidities)) + 
  geom_density(alpha = 0.5, colour = "lightblue") + theme_gray() +
  labs(x = "age (years)", title = "Distribution of cases in comorbidity groups across age", fill = "Comorbidity type") + scale_x_continuous(breaks = c(seq(18,94,4)))

```

## Cases and Controls distribution per comorbidity group

```{r, echo = FALSE, error = FALSE, message = FALSE, warning= FALSE, fig.height=4, fig.width=8, fig.align="c"}
ggplot(df, aes(x=age, colour= FinalResult)) + 
    geom_density(alpha=0.6, binwidth = 5) +
    labs(x = "age (years)", y = "density", fill = "risk factors (FR)", title = "Distribution of cases and controls across age per comorbidity") +
    theme_grey() +
    theme(legend.position = "right",
          axis.text.x = element_text(size = 6)) + facet_wrap(~comorbidities)+
  scale_x_continuous(breaks= c(seq(18, 94, 8)))
```
\newpage

## Comorbidities as risk factors in Cases

A raw account of comordities across age groups in cases is as follows. 

```{r, echo = FALSE, error = FALSE, message = FALSE, warning= FALSE, fig.align = "c"}
# Diagnosis and comorbidities
ggplot(df[df$FinalResult == "positive",], aes(x=age, fill= comorbidities)) +
    geom_histogram(alpha=0.6, binwidth = 5) +
    labs(x = "age (years)", y = "count", fill = "risk factors (FR)", title = "Counts of comorbidities in cases across age") +
    theme_grey() + 
    theme(legend.position = "right",
          axis.text.x = element_text(size = 6)) + facet_wrap(~comorbidities)+
  scale_x_continuous(breaks= c(seq(18, 94, 8)))
```

\newpage


## Comorbidities as risk factors in Controls

I compare across age groups in controls is as follows. 

```{r, echo = FALSE, error = FALSE, message = FALSE, warning= FALSE, fig.align = "c"}
ggplot(df[df$FinalResult == "negative",], aes(x=age, fill= comorbidities)) + 
    geom_histogram(alpha=0.6, binwidth = 5) +
    labs(x = "age (years)", y = "count", fill = "risk factors (FR)", title = "Counts of comorbidities in controls across age") +
    theme_grey() +
    theme(legend.position = "right",
          axis.text.x = element_text(size = 6)) + facet_wrap(~comorbidities)+
  scale_x_continuous(breaks= c(seq(4, 94, 8)))
#plot_grid(p1, p2, nrow = 1, ncol = 2)
```

\newpage
# IgG
## IgG distribution across IgG results 

Important to note are IgG ratio levels and its correspondences to IgG results regardless of final results.


```{r, echo = FALSE, error = FALSE, message = FALSE, warning= FALSE,fig.height=4, fig.width=8, fig.align="c"}

ggplot(df, aes(x = log(IgG_Ratio), fill = Sex)) + 
  geom_density(alpha = 0.5, colour = "lightblue") + theme_gray() +  theme(axis.text.x = element_text(size = 6))+
  labs(x = "IgG ratio", title = "IgG ratio per Sex") +
  scale_y_continuous(breaks = c(seq(0,6,1)))  +
   scale_x_continuous(breaks = c(seq(0,24,1)))


```

## IgG for cases and controls 

The below plot is the distribution of IgG ratio result of patients in cases and controls where there is a clearer distinction between a negative and positive final result. Positive final results correspond with higher IgG ratio levels although the peak is on the lower end on the x axis. Negative final results correspond to lower IgG ratio levels with extremely high densities.
```{r, echo = FALSE, error = FALSE, message = FALSE, warning= FALSE, fig.height=4, fig.width=8, fig.align="c"}
ggplot(df, aes(x = log(IgG_Ratio),  fill = FinalResult)) + 
  geom_density(alpha = 0.5, colour = "lightblue") + theme_gray() + theme(axis.text.x = element_text(size = 6)) +
  labs(x = "IgG_Ratio", title = "IgG ratio for cases and controls ", fill = "Final result")  +
  scale_y_continuous(breaks = c(seq(0,6,1))) +
   scale_x_continuous(breaks = c(seq(0,24,1)))
```
```{r, echo = FALSE, error = FALSE, message = FALSE, warning= FALSE,fig.height=4, fig.width=8, fig.align="c"}

# Age distribution amongst positive cases
ggplot(df) +
    geom_histogram(aes(x= log(IgG_Ratio), fill=FinalResult,
                       position = 'identity'))  + theme_gray() + labs(x = "log IgG Ratio", y = "count", fill = "Final Result") 
```

# Acknowledgements 

Synthetic data was taken from https://github.com/HopkinsIDD/serocovpop/blob/master/generated_data/serocov-pop_data_public.csv 


> I acknowledge the Traditional Aboriginal owners of land on which I find my privileges and endowments, blessings and responsibilities to my community, the Wurundjeri people of the Kulin Nations, and pay my respect to Elders past and present 


> StopAsianHate 



